---
title: "fticRms"
subtitle: "FTICR-MS data processing and visualization for petroleomics applications"
author: "Renzo C. Silva"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

## Introduction | Read me

This repository hosts a series of self-contained scripts that handle common data processing and visualization steps in ultrahigh mass spectrometry applications. In particular, Fourier transform ion cyclotron resonance mass spectrometry data applied to Petroleomics is focused herein. 

There are a few R packages able to process FTICR-MS datasets: [FTICRMS](https://cran.r-project.org/web/packages/FTICRMS/index.html) and [ftmsRanalysis](https://github.com/EMSL-Computing/ftmsRanalysis), to name a couple. Readers are encouraged to test those for a more comprehensive approach. The `fticRms` repo is not meant to become a competing package. Instead, I aim at sharing code that has been useful for a free-styler like myself.

## Input data examples

A set of 16 samples analyzed by FTICR-MS using an electrospray ion source operating at negative ion mode is provided in `/data_input/Sample*.xls`. These files are representative of an output of a commercial FTICR-MS data processing software. 

```{r echo = TRUE, message = FALSE}
library(readxl)
library(DT)
library(dplyr)
library(stringr)
library(tidyr)
library(magrittr)
library(ggplot2)

filelist <- list.files("./data_input/", full.names = TRUE)
datatable(head(read_excel(filelist[1], skip = 6),
              1000),
          options = list(scrollX = TRUE))
```

## Data clean-up and pre-processing considerations

Some data clean-up and transformation is needed at this point. A function 'pre_process' that aims at taking the dataset and selecting desired rows and columns is derived then applied to individual files in the dataset, and results are combined into a single data frame. Herein, monoisotopic intensities are selected to represent the peak intensity. 

> Note: Using monoisotopic intensities is the standard approach when representing FTICR-MS results in plots and discussions. This default option needs to be challenged when applying such data in multivariate statistical models. Should the algorithm ingest raw monoisotopic intensities? What is the best metric to ultimately associate with compound intensity? Subject matter expertise is needed here! There are dozen different approaches published in the scientific literature, but that doesn't mean they are good and robust. Discussions on the data processing strategies are beyond this walkthrough's scope. Here, for the sake of exemple, monoisotopic intensities are selected and no correction relative to baseline or monoisotopic abundance is applied.  

Different strategies for data storage optimization may be considered. Three different tables are created but they all share the same identifier based on detected molecular formulae. 

```{r echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE}

pre_process <- function(file_name) {
  # Clear out data frame
  read_excel(file_name, skip = 6) %>%
    select(c(3, 6, 15, 13, 8, 12, 16)) %>%
    filter(!is.na(Formula), DBE != "DBE")
}

# Combined filed 1st
formulae <- sapply(filelist, pre_process, simplify = FALSE) %>%
  bind_rows(.id = "Sample") %>%
  #Capture sample name from file name
  mutate(Sample = tools::file_path_sans_ext(basename(Sample))) %>%
  mutate_at(3:ncol(.), as.numeric)

# Table with intensities
intensities <- formulae %>%
  select(Sample, Formula, `Mono Inty`) %>%
  spread(key = "Sample", value = "Mono Inty")

# Table with assignment errors
errors <- formulae %>%
  select(Sample, Formula, `ppm Error`) %>%
  spread(key = "Sample", value = "ppm Error")

# Table with formula meta info
formulae <- formulae %>%
  select(Formula, `Calc m/z`, DBE, `Isotope frac`) %>%
  # Break down molecular formulas into single elements; adjust DBE values
  #A = Na, L = Cl
  mutate(
    C = str_extract(Formula, "(?<=C)[0-9]*"),
    H = str_extract(Formula, "(?<=H)[0-9]*"),
    A = str_extract(Formula, "(?<=Na)[0-9]*"),
    S = str_extract(Formula, "(?<=S)[0-9]*"),
    O = str_extract(Formula, "(?<=O)[0-9]*"),
    N = str_extract(Formula, "(?<=N)[0-9]*"),
    L = str_extract(Formula, "(?<=Cl)[0-9]*")
  ) %>%
  replace_na(list(
    A = 0,
    S = 0,
    O = 0,
    N = 0,
    L = 0
  )) %>%
  mutate_at(2:ncol(.), as.numeric) %>%
  replace_na(list(
    A = 1,
    S = 1,
    O = 1,
    N = 1,
    L = 1
  )) %>%
  mutate(Class = str_extract(Formula, "[ANSOL](.*)")) %>%
  replace_na(list(Class = "CH")) %>%
  mutate(DBE = DBE - 0.5)

```

```{r}
datatable(head(formulae,
              1000),
          options = list(scrollX = TRUE))

datatable(head(intensities,
              1000),
          options = list(scrollX = TRUE))
```

## Sample labeling

Further sample info has been provided in the input file `sampleset_info.xlsx`, which shows that 12 samples are classified as belonging to groups A|B, whereas 4 samples are labeled as "U" - unknown group. 

```{r}
labels <- read_excel("sampleset_info.xlsx")
table(labels)

```

## Data visualization

In this section, the framework to produce the most used plots for FTICR-MS analysis is presented. For such complex datasets, there are several different ways to visualize data while searching for insights that would help understand the system. A few selected plots are presented. Other supporting info derived from the original tables, such as a vector with the total sample monoisotopic intensity, are also prepared.  

```{r}
#This code chunk joins the 'labels', 'formulae' and 'intensities' tables to prepare a dataframe more adequate for the ggplot2 framework.

total_int <- intensities[,-1] %>% 
  replace(is.na(.), 0) %>%
  summarise_all(sum)




```

### Filtering the data

A filtering functionality to be applied to any plot is defined. Here, the user chooses which molecular formulae will be further considered in the plots. For example, one may want to investigate the carbon number distribution of DBE 2 species in the heteroatom class O2, and these filters will have to be dynamically adjusted. Or even if all the variables are selected, perhaps the user wants to filter out one sample from the database.  
```{r}

# User inputs required

# Define filters
select_class <- unique(formulae$Class)
select_DBE <- unique(formulae$DBE)
select_C <- unique(formulae$C)
select_samples <- unique(labels$Sample)

# Define the plot type
plot_type <- "Class" #"Class", "DBE", "C"

#Apply filters
filtered <- formulae %>%
  filter(Class %in% select_class, # Class filter
         DBE %in% select_DBE, # DBE filter
         C %in% select_C) %>% # Carbon number filter
  select(Formula, Parameter = plot_type) # Plot type filter

#Get joined data.frame for plotting
intensities %>%
  #Normalize intensities to total sample intensity
  mutate_at(vars(-Formula), funs(./sum(., na.rm = TRUE))) %>%
  gather(key = Sample, value = Intensity, 2:ncol(.)) %>%
  filter(!is.na(Intensity)) %>%
  right_join(filtered, by = "Formula") %>%
  left_join(labels, by = "Sample") %>%
  select(-Formula) %>%
  group_by(Parameter, Sample, Group) %>%
  summarise(Intensity = sum(Intensity)) %>%
  
  #Start ggplot
  ggplot(aes(
    x = Parameter,
    y = Intensity,
    fill = Sample,
    color = Group
  )) +
  geom_bar(stat = "identity", position = position_dodge()) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0, size = 14, face = "italic"),
    legend.background = element_rect(size = 1,
                                     linetype = "solid"),
    legend.position = "right"
  ) +
  labs(y= "Relative Intensity", x = "Heteroatom class") +
  scale_color_brewer(palette = "Dark2", name = "Formation") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())



```









